Demystifying the Computer Science \\ PhD Admission in the US\\{\large A Guideline for International \emph{and Domestic} Students

### 2024-03-03 demystify.tex
updates

\title{Demystifying the Computer Science \\
  PhD Admission in the US\\{\large A Handbook for International \emph{and Domestic} Students}}

@@

% though \emph{very} top schools could be very selective, e.g., see the \href{https://da-data.blogspot.com/2015/03/reflecting-on-cs-graduate-admissions.html}{admission process} at CMU
So I want to share some details about the admission process and advice for those who are interested in applying for a \textbf{PhD in Computer Science in the US}.
While this handbook is primarily intended for students interested in CS, it might be relevant to students from various disciplines.
Furthermore, although many examples are specifics for schools that I and other contributors of this document know about, the information should be generalizable to other R1\footnote{An \href{https://en.wikipedia.org/wiki/List_of_research_universities_in_the_United_States}{R1 institution} in the US is a research-intensive university with a high level of research activity across various disciplines. Currently, 146 (out of 4000) universities are classified as R1.} institutions in the US (and universities in other countries).

In addition, this handbook can help \textbf{US faculty and admission committee} gain a better understanding of international students and their cultural differences.  By recognizing and leveraging these differences, CS programs in the US can attract larger and more competitive application pools from international students.

@@

This handbook is available at

@@

@@

Additionally, \textbf{work experiences at well-known research laboratories}, such as Microsoft Research, can strengthen your
application.  Unfortunately, many good research places in your countries, e.g., VinAI in Vietnam, remain relatively unknown to most universities in the US. So you or your LoR writers should explicit say something about them in your statement.

@@

As mentioned, if you're admitted to a \emph{good} CS PhD program, you should not have to worry about funding!  
In the US, the common types of funding for PhDs are \emph{graduate teaching assistant} (GTA or TA), \emph{graduate research assistant} (GRA or RA), and \emph{Fellowship}.
RA is paid by a prof. for you to do their research. TA is paid by the department for you to help with teaching. Finally, fellowship is an independent funding that can come from a school, a company, or an organization. \autoref{tab:funding} summarizes the differences.
Note that funding is typically more available for PhD students than 
Masters.

@@

Several things about stipends.  First, the amount of stipend \emph{varies} and depends on factors such as location (e.g., a stipend in Washington DC is likely higher than in Lincoln, Nebraska due to higher living cost). Second, an academic year (AY)  year is typically \emph{9-month} in the US, so stipend is for 9 months (so divide by 9 for each month). Third, you might get paid over the summer through funding from your professor or fellowship (typically no TA over the summer). Fourth, like most source of income in the US, you will have to \emph{pay tax} on your stipend.  Finally, private universities might pay more for stipend, but they could have extra "activity" fee or some other hidden ones (e.g., you may need to pay some fees for each credit hour).

\paragraph{Low Stipend?} Students often complain their stipend being too low, but it is actually not bad and you can live comfortably with it.  In most cases, it is also enough to support your spouse and kids (many CS PhD students have their family with them). So don't worry too much about stipend.  If you're admitted to a good CS PhD program, you will be fine. A good school would know that it has to be competitive to attract students.  For example, at GMU, every year we discuss about improving the benefits, and especially stipend, for our graduate students.

@@

\begin{commentbox}[Vu:]
  At GMU CS, PhD students admitted TA has a 4 years of GTA guaranteed, and in also receive  stipend for the first summer.
\end{commentbox}

@@

\textbf{How to get RA?} When a professor recruits you, they might offer you an RA right away (e.g., when you apply).  A common scenario is that you first get admitted with TA, and then after a year or two find an adviser to support you with RA. See how to contact a prof. in \autoref{sec:contact} for research opportunities.

\begin{commentbox}[Vu:]
  If you got recruited an offered an RA by a prof., it's very likely you will get admitted.  For example, if a prof., even if not in PhD adcom, wants to fund you, the PhD adcom will respect that decision and admit you (unless your application has many red flags).
\end{commentbox}

@@

In general, fellowship, especially competitive ones from the government or company, is prestigious, and you will stand out if you get one.  Every PhD student has pubs, but only few  would have NSF Graduate or Microsoft fellowship. In fact, these are so prestigious that even if you didn't get it but make it to the final round or even ``honorable mentioning'', school will still mention you on their website and you still should put it on your CV.

@@

\section{Choosing an Adviser}
Obviously, there is no one-size-fit-all answer to this question. The best adviser is the one that you can work well with.  This is difficult and takes time. Fortunately, unlike many countries that require finding an adviser and research topic before starting the PhD (\autoref{sec:non-us-differences}), CS PhD programs in the US will typically give you a couple of years to search for advisers and research topics.  This is especially true if you're admitted with TA (\autoref{sec:ta}), which gives you time to explore and find adviser.

@@

Here are some general advice to find professor.  Assuming you're not familiar with any particular one, then first, search for profs. that share similar research interests. You can learn about faculty and their level of research activity through the faculty's website or even from CSRankings.  For example, in CSRankings, if you want to work with PL, you can search for those publish in PL conferences.  If you want to work with SE \emph{and} AI, you can search for faculty who work in both SE and AI.   You can also go to their website, look at their papers and projects.

In any case, you should contact the professor and talk to them. For example, you can (cold) email them and ask about their research and if they are taking new students (\autoref{sec:contact}), and schedule appointments to chat with them (office hrs are often the best time).

@@

Another effective way is taking graduate level courses in the topics you are interested in.  Professors teaching these \emph{special topics courses} and \emph{research seminars} might be recruiting students---giving you even a higher chance. Do well in the class, answer questions, talk to the prof after classes, etc---being stand out.  Many professors, including myself, prefer taking in new students this way.  It gives both the professor and student more time, e.g., a whole semester, to work and evaluate the relationship before making any commitment (sounds a bit like a marriage!).
You can also ask if you can do an independent study or research with a prof. This can be informal (no credit) and takes place during the summer or winter break.  For example, I do this with multiple students, many of whom are undergrads. Many will drop because they find they don't like my research, but some will stay and become my PhD students.

@@

Not every faculty can serve as your official PhD adviser. Let's try to understand the different types of faculty and their roles. For example, you might hear about tenured,  tenure track, and teaching faculty.  You might also hear about assistant, associate, full, adjunct, emeritus, teaching, research professors. Here is a primer on these terminologies.

\paragraph{Tenure-line and teaching faculty} Tenure line faculty, consisting of tenured and tenured-track profs., focus on research, which includes publishing papers, obtaining grants, and mentoring Ph.D students.  They often have low teaching load (e.g., 1 per semester). \emph{Tenured} faculty are professors who have been promoted to have a permanent position (informally, very hard to fire them).   \emph{Tenure-track} faculty are (more likely young) professors who are on the track to get tenure.  At most universities, \textbf{only} tenure and tenured track faculty can serve as a formal adviser of Ph.D. students. \autoref{sec:tenure-vs-tenure-track} talks more about choosing between tenure and tenured-track professor as your adviser.

@@

Also note that professors in a non-CS department unlikely can serve as formal PhD advisers for CS students (see \autoref{sec:related-fields}).

@@

\begin{center}
  \includegraphics[scale=0.6]{files/emails.png}
\end{center}

Faculty often receive many "cold" e-mails from prospective students. Most of the time, we ignore these emails, but in some rare occasions we do answer them. So how to write an email that get our attention?

@@

However, if your goal is a PhD in CS, then you likely need to be in the CS dept \emph{and} advised by a CS faculty. A non-CS may \emph{co-advise} or be in the PhD dissertation committee, but your main PhD adviser will likely be a tenured or tenure-track faculty in CS (\autoref{sec:faculty-types}). If in doubt, check with the CS department for their requirements.

@@

\section{Why did I get rejected?}
Many possible reasons!  Other than having weak profile (e.g., research potentials, GPA, LoRs, SOP), here are some that you might not expect:
\begin{itemize}
  \item  \emph{Your English scores (e.g., IELTS or TOEFL) are not high enough}.  While professors might not care much about these, there is a certain minimum set by the college or the university that you need to pass to get admitted, especially for TA funding (e.g., you need to be able to communicate well with students as a TA).  Moreover, while professors are willing to fight to get you admitted, they might be reluctant to go against the standard requirement of the university or college on English scores.  Note this is mostly for international students, as domestic one do not have to take these tests.

@@

\paragraph{Should I ask for feedback?}
No, don't bother.  You will likely not get any useful feedback.  We are not willing to reveal your evaluation results or give you advice on how to improve. So it is best to just move on.  If you really want feedback, ask your professors, collaborators, ref writers, or those who have previously applied.

\section{Increasing your admission chance}\label{sec:improve-your-chance}

@@

Given the high number of quality applicants and limited number of spots, other than having a good application profile, you really want to show something that makes you \textbf{stand out}, e.g., do you have a degree or background in \emph{dance} or \emph{music} and want to integrate them with CS? are you a female or a minority in CS (research for "URM minority in CS in the US" on Google)? Do you participate in outreach activities that help increase diversity and inclusion in CS? All of these are unique and would get noticed from reviewers.

@@

\chapter{Domestic Students}\label{sec:domestic-students}

Most of what is written in this document is applicable to both domestic\footnote{Domestic simply means you did your undergrad (or MS) at a US university.} and international students.  However, there are some differences that domestic students should be aware of.  This section lists some of these differences.

\section{Application Materials}

\paragraph{Fee Waiving} Some school might offer application fee waiver for domestic students.  You should check with the school you're applying to.

\paragraph{School \autoref{sec:your-school}} we already know about your school, which is a plus (see \autoref{sec:your-school}). You are also more familiar with the US education system and culture (so hopefully you are aware of the cultural issues listed in \autoref{sec:cultural}).

\paragraph{Standard Tests \autoref{sec:grades}} You do not need to take TOEFL or IELTS because you already did your undergrad (or MS).

\paragraph{Transcripts} You do not need to get your transcripts evaluated by WES or other services (which can be a hassle and inaccurate).  You can just send your official transcripts directly to the school you're applying to.

\paragraph{Funding \S\ref{sec:funding}} You have more opportunities for funding, e.g., through government scholarships for US citizens and residents.  You can also apply to specific programs \emph{before} you start your PhD, e.g., NSF Graduate Research Fellowship Program (GRFP) and Hertz Foundation Fellowship.  These fellowships can be very competitive and will improve your admission chance if you get them.

\paragraph{Research Experiences} You might have more research experiences than international students. For example, you likely have many opportunities to do research as an undergraduates, e.g., through REUs and internships.  You should highlight these experiences in your application.

\paragraph{Standing out} There are very few domestic applications compared to international ones (i.e., domestic students are minority in the CS PhD application pool).  That makes you stand out and can really help your case.

% \begin{domesticbox}[Domestic students:] If you're a domestic student, you have several advantages in your application.   You also have more opportunities for funding, e.g., through government scholarships for US citizens and residents. Finally,
% \end{domesticbox}

@@

\paragraph{History} This handbook was conceived during a lunch with Craig Yu at GMU.  We talked on about why GMU were not able to attract good Vietnamese and other international students, despite having a much stronger CS program than many schools that these students want to go to (part of the reason is described in \autoref{sec:selecting-ranking-schools}). We wished there were a way for international students to know about the US PhD programs (also for US faculty to understand more about international students and therefore have better chance of recruiting and working with them). I was also a member of the large VietPhD group on Facebook and saw many questions from students about PhD programs.  However, most active participant are students in non-CS fields or not in US, and their answers are unfortunately not always accurate and sometimes leading to more confusion. So I thought it would be useful to have a document that is specific to CS PhD programs in the US from an insider prospective.

I started writing this handbook in May 2023 and have been updating it since then (mostly around deadline time when I tend to procrastinate, i.e., \emph{productive procrastination}!). I have put the source code of this document on \href{https://github.com/nguyenthanhvuh/phd-cs-us}{GitHub} so that others can contribute to it.

@@

Though I'm not an international student, many of my students and collaborators are. I also mentor multiple students from Vietnam, and have close colleagues and friends who were once international students. I hope to capture the diverse challenges and experiences they've faced in this handbook, so that it  can be a valuable resource for prospective international students.
Finally, my upbringing in the US provides a perspective aligned with American culture, allowing me to shed light on various issues, particularly those related to cultural differences (\autoref{sec:cultural}).

\paragraph{Acknowledgement} Many people have contributed to this work.
Profs. Craig Yu (GMU), Hakan Aydin (GMU), 
Xiaokuan Zhang (GMU), Hung Le (UMass), and Deepak Kapur (UNM) provided valuable input in the early version. Other GMU faculty members also have provided feedback and contributions.  Many students including Didier (GMU), Thanh (Melbourne), and Dat (Melbourne) have contributed valuable questions and feedback.
\textbf{Thank you!}

>>>>

\title{Demystifying the Computer Science \\
  PhD Admission in the US\\{\large A Guideline for International \emph{and Domestic} Students}}

@@

% though \emph{very} top schools could be very selective, e.g., see the \href{https://da-data.blogspot.com/2015/03/reflecting-on-cs-graduate-admissions.html}{admission process} at CMU
So I want to share some details about the admission process and advice for those who are interested in applying for a \textbf{PhD in Computer Science in the US}.
While this document is primarily intended for students interested in CS, it might be relevant to students from various disciplines.
Furthermore, although many examples are specifics for schools that I and other contributors of this document know about, the information should be generalizable to other R1\footnote{An \href{https://en.wikipedia.org/wiki/List_of_research_universities_in_the_United_States}{R1 institution} in the US is a research-intensive university with a high level of research activity across various disciplines. Currently, 146 (out of 4000) universities are classified as R1.} institutions in the US (and universities in other countries).

In addition, this document can help \textbf{US faculty and admission committee} gain a better understanding of international students and their cultural differences.  By recognizing and leveraging these differences, CS programs in the US can attract larger and more competitive application pools from international students.

@@

This document is available at

@@

\begin{domesticbox}[Domestic students:] If you're a domestic student, you have several advantages in your application.  First, we already know about your school (see \autoref{sec:your-school}), and you are also more familiar with the US education system and culture.  You also have more opportunities for funding, e.g., through government scholarships for US citizens and residents. Finally, compared to international students, PhD applications in CS from domestic students are few and that can really help your case.
\end{domesticbox}

@@

Additionally, \textbf{work experiences at renowned research laboratories}, such as Microsoft Research, can strengthen your
application.  Unfortunately, many good research places in your countries, e.g., VinAI in Vietnam, remain relatively unknown to most universities in the US. So you or your LoR writers should explicit say something about them in your statement.

@@

As mentioned, if you're admitted to a \emph{good} CS PhD program, you should not have to worry about funding!  
In the US, the common types of funding for PhD are \emph{graduate teaching assistant} (GTA or TA), \emph{graduate research assistant} (GRA or RA), and \emph{Fellowship}.
RA is paid by a prof. for you to do their research. TA is paid by the department for you to help with teaching. Finally, fellowship is an independent funding that can come from a school, a company, or an organization. \autoref{tab:funding} summarizes the differences.
Note that funding is typically more available for PhD students than 
Masters.

@@

Several things about stipends.  First, the amount of stipend \emph{varies} and depends on various factors such as location (e.g., a stipend in Washington DC is likely higher than in Lincoln, Nebraska due to higher living cost). Second, an academic year (AY)  year is typically \emph{9-month} in the US, so stipend is for 9 months (so divide by 9 for each month). Third, you might get paid over the summer through funding from your professor or fellowship (typically no TA over the summer). Fourth, like most source of income in the US, you will have to \emph{pay tax} on your stipend.  Finally, private universities might pay more for stipend, but they could have higher "activity" or some other hidden fee (e.g., you may need to pay some fees for each credit hour).

Students often complain their stipend being too low, but it is actually not bad and you can live comfortably with it.  In most cases, it is also enough to support your spouse and kids (many CS PhD students have their family with them). So don't worry too much about stipend.  If you're admitted to a good CS PhD program, you will be fine. A good school would know that it has to be competitive to attract students.  For example, at GMU, every year we discuss about improving the benefits, and especially stipend, for our graduate students.

@@

\begin{commentbox}[Vu:]
  At GMU CS, students admitted with TA have  4 years of GTA guaranteed, and in some cases also receive  stipend for the first summer.
\end{commentbox}

@@

\textbf{How to get RA?} When a professor recruits you, they will likely give you RA right away (e.g., when you apply).  A common scenario is that you first get admitted with TA, and then after a year or two find an adviser to support you with RA. See how to contact a prof. in \autoref{sec:contact} for research opportunities.

\begin{commentbox}[Vu:]
  If you got recruited by a prof., who would give you RA right away, it's very likely you will get admitted.  For example, if a prof., even if not in PhD adcom, wants to work with and funds you, the PhD adcom will respect that decision and admit you (unless your application has many red flags).
\end{commentbox}

@@

In general, fellowship, especially competitive ones from the government or company, is prestigious, and you will stand out if you get one.  Every PhD student has pubs, but only superstars have NSF grad or Microsoft fellowship. In fact, these are so prestigious that even if you didn't get it but make it to the final round or even ``honorable mentioning'', school will still mention you on their website and you still should put it on your CV.

@@

\section{Choosing an Adviser}
Obviously, there is no one-size-fit-all answer to this question. The best adviser is the one that you can work well with, i.e., \emph{fits} you.  This is difficult and takes time. Fortunately, unlike many countries that require finding an adviser and research topic before starting the PhD (\autoref{sec:non-us-differences}), CS PhD programs in the US will typically give you a couple of years to search for advisers and research topics.  This is especially true if you're admitted with TA (\autoref{sec:ta}), which gives you time to explore and find adviser.

@@

Here are some general advice to find professor.  First, search for profs. that share similar research interests. You can learn about faculty and their level of research activity through the faculty's website or even from CSRankings.  For example, in CSRankings, if you want to work with PL, you can search for those publish in PL conferences.  If you want to work with SE \emph{and} AI, you can search for faculty who work in both SE and AI.  Then you can contact the professor you're interested in (\autoref{sec:contact}). %For example, you can (cold) email them and ask about their research and if they are taking new students (\autoref{sec:contact}), and schedule appointments to chat with them (office hrs are often the best time).

@@

Another effective way is taking graduate level courses in the topics you are interested in (remember: you have those 2 years to explore).  Professors teaching these \emph{special topics courses} and \emph{research seminars} might be recruiting students---giving you even a higher chance. Do well in the class, answer questions, talk to the prof after classes, etc---being stand out.  Many professors, including myself, prefer taking in new students this way.  It gives both the professor and student more time, e.g., a whole semester, to work and evaluate the relationship before making any commitment (sounds a bit like a marriage!).
You can also ask if you can do an independent study or research with a prof. This can be informal (no credit) and takes place during the summer or winter break.  For example, I do this with multiple students, many of whom are undergrads. Many will drop because they find they don't like my research, but some will stay and become my PhD students.

@@

Not every faculty can serve as your formal adviser. Let's try to understand the different types of faculty and their roles. For example, you might hear about tenured,  tenure track, and teaching faculty.  You might also hear about assistant, associate, full, adjunct, emeritus, teaching, research professors. Here is a primer on these terminologies.

\paragraph{Tenure-line and teaching faculty} Tenure line faculty, consisting of tenured and tenured-track profs., focus on research, which includes publishing papers, obtaining grants, and mentoring Ph.D students.  They often have very low teaching load (e.g., 1 per semester). \emph{Tenured} faculty are professors who have been promoted to have a permanent position (informally, very hard to fire them).   \emph{Tenure-track} faculty are (more likely young) professors who are on the track to get tenure.  \textbf{Both tenure and tenured track faculty can serve as a formal adviser of Ph.D. students}. This is very important because you need to have a tenure line faculty as your main PhD adviser. \autoref{sec:tenure-vs-tenure-track} talks more about choosing between tenure and tenured-track professor as your adviser.

@@

It is worth noting that professors in a non-CS department unlikely can serve as formal PhD advisers for CS students (see \autoref{sec:related-fields}).

@@

\begin{center}
  \includegraphics[scale=0.4]{files/emails.png}
\end{center}

Faculty received many "cold" e-mails from prospective students seeking for admission, TA, and RA. Most of the time, we ignore these emails, but in some rare occasions we do answer them. So why do we ignore some but reply to others and how to write an email that get our attention?

@@

However, if your goal is a PhD in CS, then you likely need to be in the CS dept \emph{and} advised by a CS faculty. In fact, if a faculty is not in CS, it is unlikely that they can be the \emph{main} adviser of a CS PhD student. They may \emph{co-advise} or be in the PhD dissertation committee, but your main adviser will likely be a tenured or tenure-track faculty in CS. If in doubt, check with the CS department for their requirements.

@@

\section{Why did I get rejected?}
Many possible reasons!  Other than the obvious ones (e.g., weak research pontentials, low GPA, weak LoRs, unclear SOP), here are some that you might not expect:
\begin{itemize}
  \item  \emph{Your English scores (e.g., IELTS or TOEFL) are not high enough}.  While professors do not care much about these, there is a certain minimum set by the college or the university that you need to pass to get admitted, especially if you want to get funding (e.g., you need to be able to communicate well with students as a TA).  Note this is mostly for international students, as domestic one do not have to take these tests.

@@

\paragraph{Should I ask for feedback?}
No, don't bother.  You will likely not get any useful feedback.  We cannot reveal your evaluation results (e.g., if we say because you have weak LORs, then that would reveal something about your writers) or give you advice on how to improve. So it is best to just move on.  If you really want feedback, ask your professors, collaborators, ref writers, or those who have previously applied.

\section{What can you do to increase your admission chance?}\label{sec:improve-your-chance}

@@

Show something that makes you \textbf{stand out}, e.g., do you have a degree or background in \emph{dance} or \emph{music} and want to integrate them with CS? are you a female or a minority in CS (research for "URM minority in CS in the US" on Google)? Do you participate in outreach activities that help increase diversity and inclusion in CS? All of these are unique and would get noticed from reviewers.

@@

@@

\paragraph{History} This document was conceived during a lunch with Craig Yu at GMU.  We talked on about why GMU were not able to attract good Vietnamese and other international students, despite having a much stronger CS program than many schools that these students want to go to (part of the reason is described in \autoref{sec:selecting-ranking-schools}). We wished there were a way for international students to know about the US PhD programs (also for US faculty to understand more about international students and therefore have better chance of recruiting and working with them). I was also a member of the large VietPhD group on Facebook and saw many questions from students about PhD programs.  However, most active participant are students in non-CS fields or not in US, and their answers are unfortunately not always accurate and sometimes leading to more confusion. So I thought it would be useful to have a document that is specific to CS PhD programs in the US from an insider prospective.

I started writing this document in May 2023 and have been updating it since then (mostly around deadline time when I tend to procrastinate, i.e., \emph{productive procrastination}!). I have put the source code of this document on \href{https://github.com/nguyenthanhvuh/phd-cs-us}{GitHub} so that others can contribute to it.

@@

Though I'm not an international student, many of my students and collaborators are. I also mentor multiple students from Vietnam, and have close colleagues and friends who were once international students. I hope to capture the diverse challenges and experiences they've faced in this document, so that it  can be a valuable resource for prospective international students. 
Finally, my upbringing in the US provides a perspective aligned with American culture, allowing me to shed light on various issues, particularly those related to cultural differences (\autoref{sec:cultural}).

\paragraph{Acknowledgement} Many people have contributed to this document.
Profs. Craig Yu (GMU), Hakan Aydin (GMU), 
Xiaokuan Zhang (GMU), Hung Le (UMass), and Deepak Kapur (UNM) provided valuable input in the early version. Other GMU faculty members also have provided feedback and contributions.  Many students including Didier (GMU), Thanh (Melbourne), and Dat (Melbourne) have contributed valuable questions and feedback.
\textbf{Thank you!}
